---
title: "NDVI Time Series Analysis (1984–2025)"
subtitle: "SE Arizona Study Area — Near Portal, Cochise County"
author: "Analysis Document"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
  eval: false   # Global default — individual chunks override this
---

# Overview

This document finds, downloads, and cleans NDVI (Normalized Difference Vegetation Index) data for a small study area in southeastern Arizona from the earliest reliable satellite record through 2025. **Three independent data acquisition paths are provided** — choose the one that fits your environment.

## Study Area

The bounding box is defined by two corners:

| Corner | Latitude  | Longitude   |
|--------|-----------|-------------|
| NE     | 31.939848 | −109.077395 |
| SW     | 31.935964 | −109.083181 |

The box spans roughly **580 m (E–W) × 440 m (N–S)**, placing it in the Peloncillo Mountains straddling the Arizona/New Mexico border. Vegetation in this Sky Island region transitions from Chihuahuan Desert scrub to oak woodland at higher elevations, making long-term NDVI trends ecologically informative.

## Data Source

All three options draw from **Landsat Collection 2 Level-2 Surface Reflectance** (30 m resolution), merging Landsat 4, 5, 7, 8, and 9 into a single time series. Two corrections are applied during acquisition to ensure cross-sensor consistency:

1.  **SR scaling** — Collection 2 DN values are converted to physical reflectance using the standard scale factor and offset (`reflectance = DN × 0.0000275 − 0.2`). The offset does not cancel in the NDVI formula and causes systematic underestimation if omitted.
2.  **Spectral harmonization** — L8 and L9 OLI sensors have narrower bandpasses than TM/ETM+, producing NDVI values \~0.02–0.05 lower for the same surface. Roy et al. (2016, *RSE* 185:57–70, Table 2) linear coefficients are applied to OLI red and NIR reflectance to bring them into ETM+ equivalent units before NDVI is computed.

> **Note on the 1980–1984 gap:** Landsat 5 launched March 1984. The only sensor available before that date is NOAA AVHRR (from 1981), which has 8 km resolution — a single pixel covering the entire study area. Pixel-scale NDVI is not meaningful at that resolution for a 580 × 440 m site. The practical earliest date for reliable 30 m NDVI here is therefore **1984**.

## Choosing an Acquisition Method {#sec-choose}

Three options are provided. **Activate exactly one** by setting its first chunk's `eval:` flag to `true` before rendering.

|   | Option A — GEE | Option B — Planetary Computer (Python) | Option C — Pure R |
|----|----|----|----|
| **Account needed** | Yes (free) | No | No |
| **Languages** | Python | Python | R only |
| **Processing location** | Google's servers | Local machine | Local machine |
| **Downloads rasters** | No | Yes (small clips) | Yes (small clips) |
| **Best for** | Large AOIs, long records | No-account Python workflow | R-only environment |
| **Main dependency** | `earthengine-api` | `pystac-client`, `rioxarray` | `rstac`, `gdalcubes` |

All three write identical output to `data/ndvi_raw.csv`, so the cleaning and analysis sections that follow work regardless of which path was used.

------------------------------------------------------------------------

## Language Rationale

> **Why Python for acquisition (Options A & B)?** The Google Earth Engine Python client (`earthengine-api`) is the industry-standard tool for large-scale remote sensing time series — server-side computation means no raster downloads. The Planetary Computer Python stack (`pystac-client` + `rioxarray`) is the natural Python interface to Microsoft's STAC catalog, and both ecosystems have far more documentation and community support than their R equivalents for these specific tasks. Option C demonstrates that R can do the same job, at the cost of slightly more boilerplate.

> **Why R for cleaning and analysis?** Once the data land as a CSV, R's `tidyverse` ecosystem (`dplyr`, `lubridate`, `ggplot2`) provides expressive, readable pipelines for cleaning, outlier detection, and visualisation. The `forecast` and `trend` packages add statistical time-series tools with minimal boilerplate. "Tasks that talk to the satellite archive stay in Python; tasks that make sense of the data stay in R."

------------------------------------------------------------------------

# Part 0 — Shared Constants

These coordinate constants are defined in both Python and R so each section can be run independently.

```{python}
#| label: py-constants
#| eval: false   # sourced by Options A and B — do not change

SW_LAT, SW_LON =  31.935964, -109.083181   # southwest corner (lat, lon)
NE_LAT, NE_LON =  31.939848, -109.077395   # northeast corner (lat, lon)

START_DATE = "1984-01-01"
END_DATE   = "2025-12-31"
```

```{r}
#| label: r-constants
#| eval: true   # always run — R cleaning section needs these

SW_LAT <- 31.935964; SW_LON <- -109.083181
NE_LAT <- 31.939848; NE_LON <- -109.077395

START_DATE <- "1984-01-01"
END_DATE   <- "2025-12-31"
```

------------------------------------------------------------------------

# Part 1A — Option A: Google Earth Engine (Python) {#sec-gee}

> **Requires:** A free GEE account + Google Cloud project. Run `earthengine authenticate` once in a terminal before using this section. Install: `pip install earthengine-api geemap pandas numpy`
>
> **To activate this option:** change the first chunk below from `eval: false` to `eval: true`, then render.

Computation runs entirely on Google's servers — no rasters are downloaded. This is the recommended option for reproducibility across large time ranges.

## 1A.1 Initialise Earth Engine

```{python}
#| label: gee-init
#| eval: false   # ← SET TO true TO USE THIS OPTION

import ee
import pandas as pd
import numpy as np
import os

# Replace 'your-project-id' with your GEE Cloud project ID.
# Credentials are read from the cache created by `earthengine authenticate`.
ee.Initialize(project="your-project-id")
print("Earth Engine initialised.")
```

## 1A.2 Area of Interest

```{python}
#| label: gee-aoi
#| eval: false

# GEE BBox expects (west, south, east, north) — longitude before latitude.
aoi = ee.Geometry.BBox(
    west  = SW_LON,
    south = SW_LAT,
    east  = NE_LON,
    north = NE_LAT,
)
```

## 1A.3 Cloud Masking

Landsat Collection 2 Level-2 products carry a `QA_PIXEL` band generated by the CFmask v2 (Fmask) algorithm. Each bit indicates a specific condition:

| Bit | Meaning       |
|-----|---------------|
| 1   | Dilated cloud |
| 3   | Cloud         |
| 4   | Cloud shadow  |

We retain only pixels where all three bits are zero.

```{python}
#| label: gee-cloud-mask
#| eval: false

def mask_clouds_l2(image):
    """Mask cloud/shadow pixels in a Landsat C2 SR image using QA_PIXEL bits."""
    qa = image.select("QA_PIXEL")
    clear = (
        qa.bitwiseAnd(1 << 1).eq(0)   # dilated cloud
         .And(qa.bitwiseAnd(1 << 3).eq(0))  # cloud
         .And(qa.bitwiseAnd(1 << 4).eq(0))  # cloud shadow
    )
    return image.updateMask(clear)
```

## 1A.4 NDVI Functions with SR Scaling and Harmonization

NDVI = (NIR − Red) / (NIR + Red). Two steps precede the computation:

1.  **Scale to physical reflectance** — Collection 2 DNs are stored as integers; multiply by 0.0000275 and subtract 0.2 to get values in the 0–1 range. `normalizedDifference()` is intentionally avoided here because it operates on raw DNs and the offset would introduce error.
2.  **Harmonize OLI → ETM+** (L8/L9 only) — apply Roy et al. (2016) Table 2 coefficients to the red and NIR reflectance before computing NDVI. Verify coefficients against the original paper if reusing this code.

```{python}
#| label: gee-ndvi
#| eval: false

# Collection 2 Level-2 SR scale and offset (all sensors)
C2_SCALE  = 0.0000275
C2_OFFSET = -0.2

# Roy et al. (2016) OLI → ETM+ harmonization coefficients
# Source: Remote Sensing of Environment 185:57–70, Table 2
# Direction: ETM+_reflectance ≈ slope × OLI_reflectance + intercept
HARM_RED_SLOPE, HARM_RED_INT = 0.9590, 0.0029
HARM_NIR_SLOPE, HARM_NIR_INT = 0.9828, 0.0021


def add_ndvi_tm(image):
    """NDVI for Landsat 4, 5 (TM) and 7 (ETM+) — with C2 SR scaling."""
    red = image.select("SR_B3").multiply(C2_SCALE).add(C2_OFFSET)
    nir = image.select("SR_B4").multiply(C2_SCALE).add(C2_OFFSET)
    ndvi = nir.subtract(red).divide(nir.add(red)).rename("NDVI")
    return image.addBands(ndvi)


def add_ndvi_oli(image):
    """NDVI for Landsat 8/9 (OLI/OLI-2) — C2 SR scaling + Roy et al. (2016) harmonization."""
    # Step 1: scale to physical reflectance
    red = image.select("SR_B4").multiply(C2_SCALE).add(C2_OFFSET)
    nir = image.select("SR_B5").multiply(C2_SCALE).add(C2_OFFSET)
    # Step 2: transform OLI reflectance to ETM+ equivalent
    red = red.multiply(HARM_RED_SLOPE).add(HARM_RED_INT)
    nir = nir.multiply(HARM_NIR_SLOPE).add(HARM_NIR_INT)
    # Step 3: NDVI on harmonized physical reflectance
    ndvi = nir.subtract(red).divide(nir.add(red)).rename("NDVI")
    return image.addBands(ndvi)
```

## 1A.5 Assemble Collections

```{python}
#| label: gee-collections
#| eval: false

# Landsat 4 TM (1982–1993) — sparse coverage
l4 = (ee.ImageCollection("LANDSAT/LT04/C02/T1_L2")
      .filterBounds(aoi).filterDate(START_DATE, "1993-12-31")
      .map(mask_clouds_l2).map(add_ndvi_tm))

# Landsat 5 TM (1984–2012) — primary sensor for the 1980s–2000s
l5 = (ee.ImageCollection("LANDSAT/LT05/C02/T1_L2")
      .filterBounds(aoi).filterDate("1984-01-01", "2013-06-05")
      .map(mask_clouds_l2).map(add_ndvi_tm))

# Landsat 7 ETM+ (1999–2021)
# ⚠ The Scan Line Corrector (SLC) failed 31 May 2003, leaving ~22% data gaps
# per scene. We include post-SLC scenes but flag them for the cleaning step.
l7 = (ee.ImageCollection("LANDSAT/LE07/C02/T1_L2")
      .filterBounds(aoi).filterDate("1999-01-01", "2021-12-31")
      .map(mask_clouds_l2).map(add_ndvi_tm))

# Landsat 8 OLI (2013–present)
l8 = (ee.ImageCollection("LANDSAT/LC08/C02/T1_L2")
      .filterBounds(aoi).filterDate("2013-04-11", END_DATE)
      .map(mask_clouds_l2).map(add_ndvi_oli))

# Landsat 9 OLI-2 (2021–present)
l9 = (ee.ImageCollection("LANDSAT/LC09/C02/T1_L2")
      .filterBounds(aoi).filterDate("2021-10-31", END_DATE)
      .map(mask_clouds_l2).map(add_ndvi_oli))

all_landsat = (l4.merge(l5).merge(l7).merge(l8).merge(l9)
               .select("NDVI").sort("system:time_start"))

print(f"Total scenes after cloud masking: {all_landsat.size().getInfo()}")
```

## 1A.6 Extract Time Series

```{python}
#| label: gee-extract
#| eval: false

def extract_mean_ndvi(image):
    """Reduce one scene to spatial mean NDVI over the AOI."""
    stats = image.reduceRegion(
        reducer   = ee.Reducer.mean(),
        geometry  = aoi,
        scale     = 30,
        maxPixels = 1e6,
    )
    return ee.Feature(None, {
        "date"    : image.date().format("YYYY-MM-dd"),
        "ndvi"    : stats.get("NDVI"),
        "scene_id": image.get("LANDSAT_SCENE_ID"),
        "sensor"  : image.get("SPACECRAFT_ID"),
    })

features = all_landsat.map(extract_mean_ndvi).getInfo()
records  = [f["properties"] for f in features["features"]]
df       = pd.DataFrame(records)

print(f"Rows extracted: {len(df)}")
print(df.head())
```

## 1A.7 Save to Disk

```{python}
#| label: gee-save
#| eval: false

os.makedirs("data", exist_ok=True)
df.to_csv("data/ndvi_raw.csv", index=False)
print("Saved → data/ndvi_raw.csv")
```

------------------------------------------------------------------------

# Part 1B — Option B: Microsoft Planetary Computer (Python) {#sec-pc}

> **Requires:** No account. Data is publicly accessible. Install: `pip install pystac-client planetary-computer rioxarray rasterio numpy pandas`
>
> **To activate this option:** change the first chunk below from `eval: false` to `eval: true`, then render.

Planetary Computer hosts the same Landsat Collection 2 Level-2 data as GEE. This option downloads small raster chips for the AOI and computes NDVI locally. For a 580 × 440 m clip at 30 m resolution that is roughly 20 × 15 = 300 pixels per band — negligible download size (\~50 KB per scene).

## 1B.1 Imports

```{r}
reticulate::py_require("pystac-client")    
reticulate::py_require("planetary_computer")
reticulate::py_require("rioxarray")
reticulate::py_require("tqdm")
```

```{python}
#| label: pc-imports
#| eval: true   # ← SET TO true TO USE THIS OPTION

import pystac_client
import planetary_computer
import rioxarray as rxr
import numpy as np
import pandas as pd
import os
from tqdm import tqdm   # progress bar — install with: pip install tqdm
```

## 1B.2 Query the STAC Catalog

```{python}
#| label: pc-query
#| eval: false

# Open the Planetary Computer STAC API.
# sign_inplace automatically appends a SAS token to each asset URL.
catalog = pystac_client.Client.open(
    "https://planetarycomputer.microsoft.com/api/stac/v1",
    modifier=planetary_computer.sign_inplace,
)

# Pre-filter to scenes with < 70% cloud cover reported in metadata.
# Fine-grained cloud masking happens per-pixel below.
search = catalog.search(
    collections = ["landsat-c2-l2"],
    bbox        = [SW_LON, SW_LAT, NE_LON, NE_LAT],
    datetime    = f"{START_DATE}/{END_DATE}",
    query       = {"eo:cloud_cover": {"lt": 70}},
    max_items   = None,   # retrieve all matching scenes
)

items = list(search.items())
print(f"Found {len(items)} scenes with < 70% cloud cover")
```

## 1B.3 Cloud-Mask Helper

```{python}
#| label: pc-cloud-fn
#| eval: false

def apply_cloud_mask(qa_array: np.ndarray) -> np.ndarray:
    """
    Return a boolean array (True = clear pixel) from a QA_PIXEL array.
    Identical logic to Option A — bits 1, 3, 4 indicate cloud conditions.
    """
    qa = qa_array.astype(np.int32)
    return (
        (qa & (1 << 1) == 0) &   # dilated cloud
        (qa & (1 << 3) == 0) &   # cloud
        (qa & (1 << 4) == 0)     # cloud shadow
    )
```

## 1B.4 Extract NDVI Per Scene

The TM/ETM+ vs. OLI band-naming difference means we must inspect `platform` metadata before selecting the red and NIR assets. After reading, we apply the same two corrections as Option A: (1) C2 SR scaling to physical reflectance, and (2) Roy et al. (2016) harmonization for OLI sensors.

```{python}
#| label: pc-extract
#| eval: false

# Band asset names per sensor family
BAND_MAP = {
    "landsat-4": ("SR_B3", "SR_B4"),
    "landsat-5": ("SR_B3", "SR_B4"),
    "landsat-7": ("SR_B3", "SR_B4"),   # TM/ETM+: red = B3, NIR = B4
    "landsat-8": ("SR_B4", "SR_B5"),   # OLI:     red = B4, NIR = B5
    "landsat-9": ("SR_B4", "SR_B5"),
}

OLI_SENSORS = {"landsat-8", "landsat-9"}

# Roy et al. (2016) OLI → ETM+ harmonization coefficients
# Source: Remote Sensing of Environment 185:57–70, Table 2
# Verify against original paper before reusing.
HARM_RED = (0.9590, 0.0029)   # (slope, intercept) for red band
HARM_NIR = (0.9828, 0.0021)   # (slope, intercept) for NIR band

bbox = [SW_LON, SW_LAT, NE_LON, NE_LAT]
records = []

for item in tqdm(items, desc="Processing scenes"):
    platform = item.properties.get("platform", "")
    if platform not in BAND_MAP:
        continue

    red_key, nir_key = BAND_MAP[platform]

    try:
        # Read only the small AOI window from the COG (Cloud-Optimised GeoTIFF).
        # rioxarray's clip_box issues a partial HTTP range request — it does NOT
        # download the full scene (~1 GB).
        red = rxr.open_rasterio(item.assets[red_key].href,  masked=True).squeeze()
        nir = rxr.open_rasterio(item.assets[nir_key].href,  masked=True).squeeze()
        qa  = rxr.open_rasterio(item.assets["QA_PIXEL"].href, masked=False).squeeze()

        red = red.rio.clip_box(*bbox, crs="EPSG:4326")
        nir = nir.rio.clip_box(*bbox, crs="EPSG:4326")
        qa  = qa.rio.clip_box(*bbox, crs="EPSG:4326")

        clear = apply_cloud_mask(qa.values)

        # Step 1: scale DNs to physical reflectance (C2: DN × 0.0000275 − 0.2)
        r_phys = red.values.astype(float) * 0.0000275 - 0.2
        n_phys = nir.values.astype(float) * 0.0000275 - 0.2

        # Apply cloud mask
        r_phys[~clear] = np.nan
        n_phys[~clear] = np.nan

        # Step 2: harmonize OLI → ETM+ (Roy et al. 2016, Table 2)
        if platform in OLI_SENSORS:
            r_phys = r_phys * HARM_RED[0] + HARM_RED[1]
            n_phys = n_phys * HARM_NIR[0] + HARM_NIR[1]

        # Step 3: NDVI on harmonized physical reflectance
        ndvi_mean = float(np.nanmean((n_phys - r_phys) / (n_phys + r_phys + 1e-10)))

        if not np.isnan(ndvi_mean):
            records.append({
                "date"    : item.datetime.strftime("%Y-%m-%d"),
                "ndvi"    : ndvi_mean,
                "scene_id": item.id,
                "sensor"  : platform.upper().replace("-", "_"),
            })

    except Exception as exc:
        print(f"  Skipped {item.id}: {exc}")

df = pd.DataFrame(records).sort_values("date").reset_index(drop=True)
print(f"Extracted NDVI for {len(df)} scenes")
print(df.head())
```

## 1B.5 Save to Disk

```{python}
#| label: pc-save
#| eval: false

os.makedirs("data", exist_ok=True)
df.to_csv("data/ndvi_raw.csv", index=False)
print("Saved → data/ndvi_raw.csv")
```

------------------------------------------------------------------------

# Part 1C — Option C: Pure R (`rstac` + `gdalcubes`) {#sec-r-only}

> **Requires:** No account. Data is publicly accessible. Install R packages: `install.packages(c("rstac", "gdalcubes", "sf", "terra", "tidyverse"))` GDAL ≥ 3.1 must be installed on your system (comes with most spatial R installs).
>
> **To activate this option:** change the first chunk below from `eval: false` to `eval: true`, then render.

This option demonstrates that the full acquisition pipeline is possible in R alone. It queries the same Planetary Computer STAC API and uses `gdalcubes` to stream small raster windows on demand — no Python required.

## 1C.1 Load Packages

```{r}
#| label: r-only-pkgs
#| eval: false   # ← SET TO true TO USE THIS OPTION

library(rstac)      # STAC API queries
library(gdalcubes)  # on-demand raster data cube processing
library(sf)         # spatial reference handling
library(terra)      # fallback raster I/O (used for QA masking)
library(tidyverse)
library(lubridate)

# Enable parallel reads in gdalcubes (speeds up multi-scene processing)
gdalcubes_options(parallel = TRUE)
```

## 1C.2 Query the STAC Catalog

```{r}
#| label: r-only-query
#| eval: false

# rstac connects to any STAC-compliant API.
# items_fetch() paginates automatically to retrieve all results.
items_raw <- stac("https://planetarycomputer.microsoft.com/api/stac/v1") |>
  stac_search(
    collections = "landsat-c2-l2",
    bbox        = c(SW_LON, SW_LAT, NE_LON, NE_LAT),
    datetime    = paste(START_DATE, END_DATE, sep = "/"),
    limit       = 250   # page size; items_fetch() will page through all results
  ) |>
  get_request() |>
  items_fetch()

# Sign all asset URLs with a Planetary Computer SAS token so GDAL can read them
items_signed <- items_raw |>
  items_sign(sign_fn = sign_planetary_computer())

n_items <- length(items_signed$features)
cat(sprintf("Found %d scenes\n", n_items))
```

## 1C.3 Build Sensor-Stratified Image Collections

`gdalcubes` requires knowing which asset contains each band. Because band numbers differ between TM/ETM+ (Landsat 4/5/7) and OLI (Landsat 8/9), we build two image collections and process them separately, then combine the resulting time series tables.

```{r}
#| label: r-only-collections
#| eval: false

# Helper: extract platform from a STAC item
get_platform <- function(item) item$properties$platform

# Separate by sensor family
is_tm  <- map_lgl(items_signed$features, ~ get_platform(.x) %in%
                    c("landsat-4", "landsat-5", "landsat-7"))
is_oli <- map_lgl(items_signed$features, ~ get_platform(.x) %in%
                    c("landsat-8", "landsat-9"))

features_tm  <- items_signed$features[is_tm]
features_oli <- items_signed$features[is_oli]

cat(sprintf("TM/ETM+ scenes : %d\n", length(features_tm)))
cat(sprintf("OLI scenes     : %d\n", length(features_oli)))

# Build gdalcubes image collections.
# Planetary Computer's landsat-c2-l2 STAC items use lowercase common band names
# (EO extension): "red", "nir08", "qa_pixel" — the same keys for both TM and OLI.
# Band names inside the cube match these asset keys exactly.
make_collection <- function(features) {
  if (length(features) == 0) return(NULL)
  stac_image_collection(
    features,
    asset_names = c("red", "nir08", "qa_pixel")
  )
}

col_tm  <- make_collection(features_tm)
col_oli <- make_collection(features_oli)
```

## 1C.4 Define Data Cube View

A `cube_view` specifies the target spatial extent, CRS, resolution, and time step. Setting `dt = "P1D"` gives one composite per calendar day; scenes on the same day are aggregated by the chosen `aggregation` method.

```{r}
#| label: r-only-cube-view
#| eval: false

aoi_view <- cube_view(
  srs         = "EPSG:4326",
  extent      = list(
    left   = SW_LON, right  = NE_LON,
    bottom = SW_LAT, top    = NE_LAT,
    t0     = START_DATE, t1 = END_DATE
  ),
  dx          = 0.00027,   # ≈ 30 m in decimal degrees at this latitude
  dy          = 0.00027,
  dt          = "P1D",     # one time slice per calendar day
  aggregation = "median",  # median composite when multiple scenes fall on same day
  resampling  = "bilinear"
)
```

## 1C.5 Compute NDVI and Reduce to Spatial Mean

The `apply_pixel()` expressions inline all three processing steps: cloud masking, SR scaling (DN × 0.0000275 − 0.2), and — for OLI — Roy et al. (2016) harmonization. Because TM and OLI collections are processed separately, we use two specialised helpers rather than a single generic function.

```{r}
#| label: r-only-ndvi
#| eval: false

# Cloud bits 1, 3, 4 of qa_pixel flag dilated cloud, cloud, and cloud shadow.
# Summed into a single numeric band: 0 = clear, >0 = cloudy.
# This avoids logical operators (&&/and) which are not reliably supported
# across gdalcubes versions — only arithmetic and a single == 0 test needed.
CLOUD_BITS <- "floor(qa_pixel/2)%2 + floor(qa_pixel/8)%2 + floor(qa_pixel/16)%2"


# ── TM / ETM+ (L4, L5, L7): scale only, no harmonization needed.
# PC asset names: "red" = SR_B3, "nir08" = SR_B4
process_tm <- function(col, sensor_label) {
  if (is.null(col)) return(tibble())

  raster_cube(col, aoi_view) |>
    apply_pixel(
      c(
        "red * 0.0000275 - 0.2",
        "nir08 * 0.0000275 - 0.2",
        CLOUD_BITS
      ),
      names = c("red_phys", "nir_phys", "cloud_bits")
    ) |>
    apply_pixel(
      "cloud_bits == 0 ? (nir_phys - red_phys) / (nir_phys + red_phys) : NAN",
      names = "ndvi"
    ) |>
    reduce_space("mean(ndvi)") |>
    as_tibble() |>
    rename(date = time, ndvi = ndvi_mean) |>
    filter(!is.na(ndvi)) |>
    mutate(date = as.Date(date), sensor = sensor_label)
}


# ── OLI (L8, L9): scale + Roy et al. (2016) harmonization.
# Source: Remote Sensing of Environment 185:57–70, Table 2
# PC asset names: "red" = SR_B4, "nir08" = SR_B5
process_oli <- function(col, sensor_label) {
  if (is.null(col)) return(tibble())

  raster_cube(col, aoi_view) |>
    apply_pixel(
      c(
        "0.9590 * (red * 0.0000275 - 0.2) + 0.0029",
        "0.9828 * (nir08 * 0.0000275 - 0.2) + 0.0021",
        CLOUD_BITS
      ),
      names = c("red_harm", "nir_harm", "cloud_bits")
    ) |>
    apply_pixel(
      "cloud_bits == 0 ? (nir_harm - red_harm) / (nir_harm + red_harm) : NAN",
      names = "ndvi"
    ) |>
    reduce_space("mean(ndvi)") |>
    as_tibble() |>
    rename(date = time, ndvi = ndvi_mean) |>
    filter(!is.na(ndvi)) |>
    mutate(date = as.Date(date), sensor = sensor_label)
}

ts_tm  <- process_tm(col_tm,   "LANDSAT_TM_ETM")
ts_oli <- process_oli(col_oli, "LANDSAT_OLI")

df_r <- bind_rows(ts_tm, ts_oli) |>
  arrange(date) |>
  # gdalcubes does not expose scene-level IDs; use date+sensor as surrogate
  mutate(scene_id = paste(date, sensor, sep = "_"))

cat(sprintf("Extracted %d observations\n", nrow(df_r)))
glimpse(df_r)
```

## 1C.6 Save to Disk

```{r}
#| label: r-only-save
#| eval: false

dir.create("data", showWarnings = FALSE)
write_csv(df_r, "data/ndvi_raw.csv")
cat("Saved → data/ndvi_raw.csv\n")
```

------------------------------------------------------------------------

# Part 2 — Data Cleaning (R)

This section reads `data/ndvi_raw.csv` regardless of which acquisition option was used. All three options write the same four columns: `date`, `ndvi`, `scene_id`, `sensor`.

## 2.1 Load Packages

```{r}
#| label: r-clean-pkgs
#| eval: true

library(tidyverse)
library(lubridate)
library(zoo)
library(knitr)
```

## 2.2 Load and Inspect Raw Data

```{r}
#| label: r-load
#| eval: true

raw <- read_csv(
  "data/ndvi_raw.csv",
  col_types = cols(
    date     = col_date(format = "%Y-%m-%d"),
    ndvi     = col_double(),
    scene_id = col_character(),
    sensor   = col_character()
  )
)

glimpse(raw)

raw |>
  count(sensor, name = "n_scenes") |>
  kable(caption = "Scene count by sensor (raw)")

cat(sprintf(
  "Date range: %s  →  %s\n",
  min(raw$date, na.rm = TRUE),
  max(raw$date, na.rm = TRUE)
))
```

```{r}
#| label: fig-raw-ts
#| eval: true
#| fig-cap: "Raw NDVI values before cleaning, coloured by sensor."

ggplot(raw, aes(x = date, y = ndvi, colour = sensor)) +
  geom_point(alpha = 0.35, size = 0.7) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Raw Landsat NDVI — all scenes",
       x = NULL, y = "Mean NDVI", colour = "Sensor") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 2.3 Cleaning Pipeline

### Step 1 — Drop rows with no valid pixels

Some scenes return `NA` if the AOI was 100% cloud-covered after masking.

```{r}
#| label: r-step1
#| eval: true

step1 <- raw |> filter(!is.na(ndvi))
cat(sprintf("Step 1 — dropped %d NA rows\n", nrow(raw) - nrow(step1)))
```

### Step 2 — Physical bounds filter

NDVI is bounded \[−1, 1\] by definition. Values below −0.2 or above 0.9 for a semi-arid shrubland almost always indicate residual cloud contamination, instrument artefact, or snow/ice.

```{r}
#| label: r-step2
#| eval: true

NDVI_MIN <- -0.2
NDVI_MAX <-  0.9

step2 <- step1 |> filter(between(ndvi, NDVI_MIN, NDVI_MAX))
cat(sprintf(
  "Step 2 — removed %d rows outside [%.1f, %.1f]\n",
  nrow(step1) - nrow(step2), NDVI_MIN, NDVI_MAX
))
```

### Step 3 — Flag Landsat 7 SLC-off scenes

The Scan Line Corrector failed on 31 May 2003. Post-failure scenes have \~22% striped data loss. For a small AOI the spatial mean may be biased if the missing stripes happen to fall on vegetation. We flag rather than remove so analysts can choose to exclude them.

```{r}
#| label: r-step3
#| eval: true

SLC_FAIL <- as.Date("2003-06-01")

step3 <- step2 |>
  mutate(slc_off = (sensor %in% c("LANDSAT_7", "LANDSAT_TM_ETM")) &
                    (date >= SLC_FAIL))

cat(sprintf(
  "Step 3 — flagged %d Landsat 7 SLC-off scenes (%.1f%%)\n",
  sum(step3$slc_off),
  100 * mean(step3$slc_off)
))
```

### Step 4 — Month-stratified outlier removal

Residual cloud contamination typically drives NDVI unusually low; smoke can drive it high. A 3×IQR fence computed within each calendar month avoids discarding the legitimately high NDVI values that occur during the summer monsoon (July–September) — a global fence would treat monsoon-season peaks as outliers.

```{r}
#| label: r-step4
#| eval: true

step4 <- step3 |>
  mutate(month = month(date)) |>
  group_by(month) |>
  mutate(
    q25     = quantile(ndvi, 0.25, na.rm = TRUE),
    q75     = quantile(ndvi, 0.75, na.rm = TRUE),
    iqr     = q75 - q25,
    outlier = ndvi < (q25 - 3 * iqr) | ndvi > (q75 + 3 * iqr)
  ) |>
  ungroup() |>
  filter(!outlier) |>
  select(-q25, -q75, -iqr, -outlier, -month)

cat(sprintf(
  "Step 4 — removed %d monthly outliers\n",
  nrow(step3) - nrow(step4)
))
```

### Step 5 — De-duplicate same-date observations

Path/row overlaps can produce two scenes on the same calendar day. We keep the observation with the higher NDVI (a proxy for less cloud contamination when values are otherwise similar).

```{r}
#| label: r-step5
#| eval: true

step5 <- step4 |>
  group_by(date) |>
  slice_max(ndvi, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

cat(sprintf(
  "Step 5 — de-duplicated: %d → %d rows\n",
  nrow(step4), nrow(step5)
))
```

## 2.4 Cleaning Summary

```{r}
#| label: r-clean-summary
#| eval: true

tibble(
  Step = c(
    "Raw", "1 — NA removal", "2 — range filter",
    "3 — SLC flag", "4 — outlier removal", "5 — de-duplication"
  ),
  `Row count`    = c(nrow(raw), nrow(step1), nrow(step2),
                     nrow(step3), nrow(step4), nrow(step5)),
) |>
  mutate(`Rows removed` = lag(`Row count`, default = nrow(raw)) - `Row count`) |>
  kable(caption = "Data volume at each cleaning step")
```

## 2.5 Save Cleaned Data

```{r}
#| label: r-save-clean
#| eval: true

dir.create("data", showWarnings = FALSE)
write_csv(step5, "data/ndvi_clean.csv")
cat("Saved → data/ndvi_clean.csv\n")
```

------------------------------------------------------------------------

# Part 3 — Analysis & Visualisation (R)

## 3.0 Cross-Sensor Consistency Check

Before interpreting trends, verify that harmonization has reduced the systematic NDVI offset between L7 ETM+ and L8 OLI. The two sensors operated simultaneously from April 2013 to December 2021, giving a direct overlap window for comparison. After correct scaling and harmonization, the monthly mean NDVI distributions for L7 and L8 should be indistinguishable. A persistent offset would indicate that the coefficients need review.

```{r}
#| label: r-load-clean
#| eval: true

library(forecast)  # STL decomposition
library(ggridges)  # ridge plots
library(patchwork) # combine panels
library(scales)

ndvi_clean <- read_csv("data/ndvi_clean.csv",
                       col_types = cols(date = col_date()))
```

```{r}
#| label: fig-sensor-overlap
#| eval: true
#| fig-cap: "Monthly mean NDVI for L7 and L8 during their overlap period (2013–2021). After harmonization the distributions should be centred at similar values."

# L7/L8 overlap window
OVERLAP_START <- as.Date("2013-04-11")
OVERLAP_END   <- as.Date("2021-12-31")

# Identify the sensor column values for L7 and L8 regardless of which
# acquisition option was used (sensor names differ slightly between options)
overlap <- ndvi_clean |>
  filter(date >= OVERLAP_START, date <= OVERLAP_END) |>
  filter(grepl("7|ETM", sensor, ignore.case = TRUE) |
         grepl("8|OLI",  sensor, ignore.case = TRUE)) |>
  mutate(sensor_label = case_when(
    grepl("7|ETM", sensor, ignore.case = TRUE) ~ "Landsat 7 ETM+",
    grepl("8|OLI", sensor, ignore.case = TRUE) ~ "Landsat 8 OLI",
    TRUE ~ sensor
  )) |>
  mutate(ym = floor_date(date, "month")) |>
  group_by(ym, sensor_label) |>
  summarise(ndvi = mean(ndvi, na.rm = TRUE), .groups = "drop")

p_ts <- ggplot(overlap, aes(x = ym, y = ndvi, colour = sensor_label)) +
  geom_line(alpha = 0.7, linewidth = 0.6) +
  geom_smooth(se = FALSE, linewidth = 1.2, method = "loess", span = 0.3) +
  scale_colour_manual(values = c("Landsat 7 ETM+" = "#e41a1c",
                                  "Landsat 8 OLI"  = "#377eb8")) +
  labs(title = "L7 vs L8 monthly mean NDVI during overlap (2013–2021)",
       subtitle = "Indistinguishable lines indicate successful harmonization",
       x = NULL, y = "Monthly mean NDVI", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_box <- ggplot(overlap, aes(x = sensor_label, y = ndvi, fill = sensor_label)) +
  geom_boxplot(alpha = 0.6, outlier.size = 0.8) +
  scale_fill_manual(values = c("Landsat 7 ETM+" = "#e41a1c",
                                "Landsat 8 OLI"  = "#377eb8")) +
  labs(x = NULL, y = "Monthly mean NDVI") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

p_ts + p_box + plot_layout(widths = c(3, 1))
```

## 3.1 Long-Term Time Series

```{r}
#| label: fig-clean-ts
#| eval: true
#| fig-cap: "Cleaned NDVI time series with a 1-year rolling mean."

ndvi_roll <- ndvi_clean |>
  arrange(date) |>
  mutate(
    ndvi_roll = rollapply(ndvi, width = 23, FUN = mean,
                          fill = NA, align = "center", partial = TRUE)
  )

ggplot(ndvi_roll, aes(x = date)) +
  geom_point(aes(y = ndvi, colour = sensor), alpha = 0.25, size = 0.6) +
  geom_line(aes(y = ndvi_roll), colour = "black", linewidth = 0.8) +
  scale_colour_brewer(palette = "Set1", name = "Sensor") +
  scale_x_date(breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "Landsat NDVI — SE Arizona (1984–2025)",
    subtitle = "Points: individual scenes  |  Line: 1-year rolling mean",
    x = NULL, y = "Mean NDVI"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 3.2 Seasonal Patterns by Decade

The North American Monsoon drives a predictable summer green-up (July–September). Ridge plots reveal whether the seasonal amplitude or timing has shifted.

```{r}
#| label: fig-seasonal-ridge
#| eval: true
#| fig-cap: "Monthly NDVI distributions by decade. Rightward shifts indicate higher vegetation productivity."

ndvi_clean |>
  mutate(
    month_name = factor(month.abb[month(date)], levels = month.abb),
    decade     = paste0(floor(year(date) / 10) * 10, "s")
  ) |>
  ggplot(aes(x = ndvi, y = month_name, fill = decade)) +
  geom_density_ridges(alpha = 0.55, scale = 1.2, rel_min_height = 0.02) +
  scale_fill_viridis_d(option = "plasma", direction = -1) +
  labs(title = "Monthly NDVI distributions by decade",
       x = "Mean NDVI", y = NULL, fill = "Decade") +
  theme_ridges(font_size = 11) +
  theme(legend.position = "right")
```

## 3.3 Annual Monsoon-Season NDVI

July–September (JAS) NDVI is a standard productivity index for Chihuahuan / Sonoran Desert grasslands and the most ecologically meaningful seasonal window in this region.

```{r}
#| label: fig-annual-monsoon
#| eval: true
#| fig-cap: "Annual monsoon-season (JAS) mean NDVI with a linear OLS trend."

monsoon_annual <- ndvi_clean |>
  filter(month(date) %in% 7:9) |>
  group_by(year = year(date)) |>
  summarise(ndvi_mean = mean(ndvi, na.rm = TRUE), n = n(), .groups = "drop") |>
  filter(n >= 3)   # need ≥ 3 scenes to compute a stable seasonal mean

ggplot(monsoon_annual, aes(x = year, y = ndvi_mean)) +
  geom_col(fill = "#4dac26", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(1985, 2025, 5)) +
  labs(
    title    = "Annual monsoon-season (JAS) mean NDVI",
    subtitle = "Bar = annual mean  |  Line = OLS trend ± 95% CI",
    x = "Year", y = "Mean NDVI (JAS)"
  ) +
  theme_minimal(base_size = 12)
```

## 3.4 Long-Term Trend Test (Mann–Kendall)

The Mann–Kendall test is non-parametric and robust to non-normal distributions and tied values — appropriate for NDVI data. We apply it to the annual monsoon means (one value per year) to avoid autocorrelation inflating significance in the scene-level series.

```{r}
#| label: r-mk-test
#| eval: true

if (!requireNamespace("trend", quietly = TRUE)) install.packages("trend")
library(trend)

mk <- mk.test(monsoon_annual$ndvi_mean)

cat("Mann–Kendall trend test on annual monsoon (JAS) NDVI\n")
cat(sprintf("  tau        = %+.3f\n", mk$estimates["tau"]))
cat(sprintf("  p-value    = %.4f\n",  mk$p.value))
cat(sprintf("  Direction  : %s\n",
    ifelse(mk$estimates["tau"] > 0, "increasing", "decreasing")))
cat(sprintf("  Significant at α=0.05: %s\n",
    ifelse(mk$p.value < 0.05, "YES", "no")))
```

## 3.5 STL Decomposition

STL (Seasonal-Trend decomposition using Loess) separates the monthly mean NDVI into trend, seasonal, and remainder components. Short gaps (≤ 2 months) are interpolated before decomposition.

```{r}
#| label: fig-stl
#| eval: true
#| fig-cap: "STL decomposition of monthly mean NDVI into trend, seasonal cycle, and residuals."

# Aggregate to monthly means
monthly <- ndvi_clean |>
  mutate(ym = floor_date(date, "month")) |>
  group_by(ym) |>
  summarise(ndvi = mean(ndvi, na.rm = TRUE), .groups = "drop") |>
  complete(ym = seq(min(ym), max(ym), by = "month")) |>
  arrange(ym) |>
  mutate(ndvi_interp = na.approx(ndvi, maxgap = 2, na.rm = FALSE))

# Build regular ts object
ndvi_ts <- ts(
  monthly$ndvi_interp,
  start     = c(year(min(monthly$ym)), month(min(monthly$ym))),
  frequency = 12
)

stl_fit <- stl(ndvi_ts, s.window = "periodic", robust = TRUE,
               na.action = na.interp)

autoplot(stl_fit) +
  labs(
    title   = "STL decomposition of monthly mean NDVI",
    caption = "s.window = periodic  |  robust = TRUE"
  ) +
  theme_minimal(base_size = 11)
```

------------------------------------------------------------------------

# Session Information

```{r}
#| label: r-session
#| eval: true
#| code-fold: true

sessionInfo()
```
